Fail Faster

### 2025-03-06 paper/paper.tex
fix lstlisting
work

\lstset{
  basicstyle=\small\fontfamily{cmtt}\selectfont, % Computer Modern Typewriter
}

@@

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, splice, stage distinction.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
  \end{itemize}
}

@@

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show allocation benchmarks of the running example, versus the version where you inline everything.
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}

% In OCaml, garbage collection is the killer.

% This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
% 3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.

% \jwc{Mention ppx_jane and let syntax}

% \begin{lstlisting}
% let gen_list_bool =
%   fixed_point (fun gl ->
%     let%bind n = size in
%     if n <= 0 then return [] else
%       let%bind x = bool in
%       let%bind xs = with_size ~size:(n-1) gl in
%       return (x :: xs)
%   )
% \end{lstlisting}

% On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
% \begin{lstlisting}
%   let rec gen_list_bool_faster ~size ~random  = 
%     if size <= 0 then []
%     else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
% \end{lstlisting}

@@

\jwc{
  \begin{itemize}
    \item Naked binds like \texttt{Bind}
  \end{itemize}
}

@@

\section{\name: A Library for Staging Generators}

@@

\jwc{
  \begin{itemize}
    \item The simplest combinators are presented below.
    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
    \texttt{code} for the result of running the first generator. At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
  \end{itemize}
}

@@

  let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
    fun size_c random_c ->
      .<
        let x = .~(g size_c random_c) in
        .~(
          (k .<x>.) size_c random_c
        )
      >.

  let int (lo : int code) (hi : int code) : int t =
    fun size_c random_c ->
      .< SR.int .~random_c .~lo .~hi >.
end
\end{lstlisting}

@@

\begin{lstlisting}
let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
  .<
    Bq.create (fun size random ->
      .~(g .<size>. .<random>.)
    )
  >.
\end{lstlisting}

@@

\begin{lstlisting}
let int_pair_staged : ((int * int) code) Gen.t =
  Gen.bind (int .<0>. .<100>.) (fun cx ->
    Gen.bind (int .<0> cx) (fun cy ->
      Gen.return .<(.~cx,.~cy)>.
    )
  )
let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
====
.<
  Bq.create (fun size random ->
    let x = SR.int random 0 100 in
    let y = SR.int random 0 x in
    (x,y)
  )
.>
\end{lstlisting}

\jwc{So if you write with the \texttt{Gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}

\jwc{Show benchmark difference between these two generators.}

@@

\subsection{A Real Monad Instance, CPS, and ``The Trick''}

@@

  let return (x : 'a) : 'a t = fun _ _ -> x

>>>>

\lstset{
  mathescape=true,
  frame=none,
  xleftmargin=10pt,
  stepnumber=1,
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  % language=haskell,
  keepspaces=true,
  tabsize=2,
  emphstyle={\bf},
  % commentstyle=\it\color{dkgreen},
  stringstyle=\mdseries\ttfamily,
  showspaces=false,
  keywordstyle=\bfseries\ttfamily,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  showstringspaces=false,
  % morecomment=[l]\%,
  % moredelim=**[is][\color{dkgreen}]{@}{@}
}

@@

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, splice, stage distinction.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
    \item We can convert one way, but not the other.
    \item Functions \texttt{'a code -> 'b code} ``fuse''.
    \begin{itemize}
      \item Consider writing \texttt{even . succ}. This
      \item Consider \texttt{even_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ_c : int code -> int code = fun cx -> .< .~x + 1 >.}
      \item If you do \texttt{to_dyn (even_c . succ_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
      a dynamic function value ensures that the resulting code is free of abstractions.
    \end{itemize}
    and 
  \end{itemize}
}

@@

\jwc{We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}

\jwc{
  \begin{itemize}
    \item How does (simplified) BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show benchmarks of the running example, versus the version where you inline everything. ()
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}

\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  Bq.bind (Bq.gen_int 0 100) (fun x ->
    Bq.bind (Bq.gen_int 0 x) (fun y ->
      Bq.return (x,y)
  ))

let int_pair_inlined : (int * int) Bq.t
  fun sr ->
    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
    let y = Splittable_random.int sr ~lo:0 ~hi:x in
    (x,y)
\end{lstlisting}

\jwc{Show benchmark difference between these two generators: the benchmark is in \texttt{waffle-house/handwritten-ocaml/bin/basic-compare.ml}. It's about 2x. (see comment in latex here.)}

\jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}

% ┌───────┬──────────┬─────────┬────────────┐
% │ Name  │ Time/Run │ mWd/Run │ Percentage │
% ├───────┼──────────┼─────────┼────────────┤
% │ bq    │  70.47ns │  78.00w │    100.00% │
% │ fused │  35.86ns │  78.00w │     50.89% │
% └───────┴──────────┴─────────┴────────────┘

\jwc{The overhead of the abstraction is 2x.
The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}

@@

\jwc{
  \begin{itemize}
    \item In cases where the compiler cannot statically eliminate them, binds allocate short-lived closures: we allocate a closure for the continuation, and then immediately jump into it.
    \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast, and each minor allocation brings us closer to the next costly GC pause.
  \end{itemize}
}

@@

\section{\name: A Library for Staged Generators}

@@

\jwc{
  \begin{itemize}
    \item The simplest combinators are presented below.
    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
    \texttt{code} for the result of running the first generator. At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
    \item Bind takes code for the size and seed, and returns code that first let-binds the result of the first generator.
    The function application \texttt{g size_c random_c} runs during the first stage, producing the code for the \texttt{a}. This is then spliced into the quote: the specialized code of the generator \texttt{g}
    put on the right-hand side of the let. Then, we splice in the code for the continuation, passing first the variable x, and the same size and random seed.
    \item \texttt{}
  \end{itemize}
}

@@

  let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
    fun size_c random_c ->
      .<
        let x = .~(g size_c random_c) in
        .~(k .<x>. size_c random_c)
      >.

  let int (lo : int code) (hi : int code) : int t =
    fun size_c random_c ->
      .< SR.int .~random_c .~lo .~hi >.

  let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
  .<
    fun size random -> .~(g .<size>. .<random>.)
  >.
end
\end{lstlisting}

@@

@@

\begin{lstlisting}
let int_pair_staged : ((int * int) code) Gen.t =
  Gen.bind (Gen.int .<0>. .<100>.) (fun cx ->
    Gen.bind (Gen.int .<0> cx) (fun cy ->
      Gen.return .<(.~cx,.~cy)>.
    )
  )
let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
====
.<
  fun size random ->
    let x = SR.int random 0 100 in
    let y = SR.int random 0 x in
    (x,y)
.>
\end{lstlisting}

\jwc{So if you write with the \texttt{Gen} combinators and then only call \texttt{to\_bq} at the end, you get something that inlines everything all the way down!}

@@

\subsection{Size Parameters, A Real Monad Instance, CPS, and ``The Trick''}

\jwc{Introduce the size parameter here. (if you deleted it beforehand.)}

@@

  let return (x : 'a) : 'a t = fun _ _ -> Codecps.return x
